---
title: "Exercises chapter 10"
author: "Andreas Slåttelid"
header-includes: \usepackage{bbm}
output:
  pdf_document:
    extra_dependencies:
    - bbm
    - threeparttable
    - dsfont
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 


## Exercise 10.2 

### i)
We are asked to show that the solution to the GBM under $Q$ is given by: 
$$\begin{aligned}
S(t) &= S(0)\exp\left(
(r-\frac{\sigma^{2}}{2})t
+ \sigma \widetilde{W}_{t}
\right)
\end{aligned}$$

Now from arbitrage theory, we have that $\widetilde{S}_{t}$ is a $(Q,\mathcal{F})$-martingale, \
here I just state the dynamics of $\widetilde{S}$ under $P$: 

$$
\begin{aligned}
d(e^{-rt}S_{t}) &= 
S_{t}e^{-rt}\sigma\left[
\frac{\mu - r}{\sigma}dt + dW_{t}
\right] \\ 
&= 
\widetilde{S}_{t}\sigma[\theta dt + dW_{t}]
\end{aligned}$$

From Girsanov's thm we have that $Z_{t}$ a martingale 
$$\begin{aligned}
Z_{t} &= \exp\left(
\int_{0}^{t}\varphi_{s}dW_{s} 
-\frac{1}{2}\int_{0}^{t}\varphi_{s}^{2}ds
\right)
\end{aligned}
$$ 

yields that $\widetilde{W}_{t} = W_{t} - \int_{0}^{t}\varphi_{s}ds$ is a $(Q,\mathcal{F})$-BM. 

So combining the martingale representation theorem with a clever way of finding $\varphi_{s}$, we see that $\varphi_{t} = -\theta$, because, we then get that: 
$$
\widetilde{W}_{t} = W_{t}+\int_{0}^{t}\theta ds
$$
is a $(Q,\mathcal{F})$-BM. This gives us $W_{t} = \widetilde{W}_{t} - \theta t$, from theory we have that the solution to a GBM under $(P)$ is:  
$$\begin{aligned}
S(t) &= S(0)\exp\left(
\left(\mu - \frac{1}{2}\sigma^{2} \right)t + \sigma W_{t}
\right) \;\;(P) \\ 
&= S(0)\exp\left(
\left(\mu - \frac{1}{2}\sigma^{2} \right)t + \sigma (\widetilde{W}_{t} - \theta t)
\right) \;\;(Q) \\ 
&= S(0)\exp\left(
\left(\mu - \frac{\sigma^{2}}{2} -\sigma \theta \right)t + \sigma \widetilde{W}_{t}
\right) \\ 
&= S(0)\exp\left(
\left(r-\frac{\sigma^{2}}{2}\right)t
+ \sigma \widetilde{W}_{t}
\right)
\end{aligned}$$

\newpage 
### ii) 
We are asked to find the price of the claim $X = \max(NS_{T},G)$, now from mathematical theory we have that the price is given by: 
$$
\pi(t) = e^{-r(T-t)}\mathbb{E}_{Q}[X|\mathcal{F}_{t}]
$$
The goal here is to use the explicit formula given by Black & Scholes: 
$$
BS(t,T,S_{t}, K) = e^{-r(T-t)}\mathbb{E}_{Q}[(S_{T}-K)^{+}|\mathcal{F}_{t}]
$$

where $(x)^{+} = \max(x,0)$, so let's do so, before we dig in, we show some useful tricks concerning the maximum: 
$$\begin{aligned}
\max(A,B) &= \max(B-A,0) + A \\
\max(A,B) &= \max(A-B,0) + B \\ 
\max(A,B) &= (A-B)^{+} + B
\end{aligned}$$
Hence: 


$$\begin{aligned}
\max(NS_{T},G) &= (NS_{T}-G)^{+} + G \\ 
&\Downarrow \\ 
\frac{\max(NS_{T},G)}{N} &= \left(S_{T}-\frac{G}{N}\right)^{+} + \frac{G}{N} \\ 
&= Z + \frac{G}{N} \\ 
&\Downarrow \\ 
\max(NS_{T},G) &= NZ + G
\end{aligned}$$

Now let's use finance theory: 
$$\begin{aligned}
\pi(t) &= e^{-r(T-t)}\mathbb{E}_{Q}[X|\mathcal{F}_{t}] \\ 
&= e^{-r(T-t)}\mathbb{E}_{Q}[( NZ + G)|\mathcal{F}_{t}] \\ 
&= Ne^{-r(T-t)}\mathbb{E}_{Q}[Z|\mathcal{F}_{t}] + e^{-r(T-t)}G \\ 
&= Ne^{-r(T-t)}\mathbb{E}_{Q}\left[\left(S_{T}-\frac{G}{N}\right)^{+}|\mathcal{F}_{t}\right] + e^{-r(T-t)}G \\ 
&= N BS(t,T,S_{t}, G/N) + e^{-r(T-t)}G
\end{aligned}$$

\newpage 

## Exercise 10.3 

We are dealing with an endownment insurance, hence $S = \{*,\dagger\}$, the contractual information provided is: 
```{r}
T     <- 10    #length of contract
x0    <- 60    #age of insured
pi0   <- 10000 
S0    <- 1
beta  <- 0.005 #deduction charges
r     <- 0.025 
sigma <- 0.25  #volatility
```

From the text, we get: 
$$\begin{aligned}
C(T) &= \max((1-0.03)\pi_{0}S_{T}(1-\beta)^{T-1}, \pi_{0}) \\
C(T) &= \max(NS_{T},\pi_{0}) \\ 
C(T) &= N\left(S_{T}-\frac{\pi_{0}}{N}\right)^{+} + \pi_{0}
\end{aligned}$$

We also write down the policy functions, where we do not include premiums: 
$$\begin{aligned}
a_{*}(t) &= 
\begin{cases}
0, &t<0 \\
0, &t\in[0,T) \\ 
C(T), &t\geq T
\end{cases}\;\; 
a_{\dagger}(t) = 0 \;\; 
a_{*\dagger}(t) = 0 \;\; 
\dot{a}_{*}(t) = 0
\end{aligned}$$

We also recall what we mean by the B&S-notation: 
$$\begin{aligned}
BS(t,T,S_{t}, K) &= e^{-r(T-t)}\mathbb{E}_{Q}\left[
\left(S_{T}-K\right)^{+}
\bigg{|}\mathcal{F}_{t}
\right]
\end{aligned}$$


This yields: 
$$\begin{aligned}
V_{*}^{+}(t,S_{t}) &= p_{**}^{x_{0}}(t,T)\frac{v(T)}{v(t)}\mathbb{E}_{Q}[C(T)|\mathcal{F}_{t}] \\ 
&=  p_{**}^{x_{0}}(t,T)e^{-r(T-t)}\mathbb{E}_{Q}[N\left(S_{T}-\frac{\pi_{0}}{N}\right)^{+} + \pi_{0}|\mathcal{F}_{t}] \\ 
&=
p_{**}^{x_{0}}(t,T)Ne^{-r(T-t)}\mathbb{E}_{Q}\left[N\left(S_{T}-\frac{\pi_{0}}{N}\right)^{+}|\mathcal{F}_{t}\right] +  p_{**}^{x_{0}}(t,T)e^{-r(T-t)}\pi_{0} \\ 
&= p_{**}^{x_{0}}(t,T)N BS(t,T,S_{t}, \pi_{0}/N) + p_{**}^{x_{0}}(t,T)e^{-r(T-t)}\pi_{0} \\ 
&= p_{**}^{x_{0}}(t,T)\left[
N BS(t,T,S_{t}, \pi_{0}/N) + e^{-r(T-t)}\pi_{0}
\right]
\end{aligned}$$

\newpage

```{r}
#0: alive, 1:dead
mu01 <- function(t){
  A <- 0.0001
  B <- 0.00035
  c <- 1.075
  return(A + B*c^{t})
}

#survival function:
p_surv <- function(t,s){
  f <- mu01
  integral <- integrate(f, lower = t, upper = s)$value
  return(exp((-1)*integral))
}

p_surv(60,70)
#Black&Scholes for call-option:
BS <- function(t,T,St, K){
   d1 <- (log(St/K) + (r + sigma^2/2)*(T-t)) / (sigma*sqrt(T-t))
   d2 <- d1 - sigma*sqrt(T-t)
   
   value <- St*pnorm(d1) - K*exp(-r*(T-t))*pnorm(d2)
   return(value)
}

V_reserve <- function(t, T,St){
  
  N <- (1-0.03)*pi0*(1-beta)^{(T-1)}
  
  reserve <- p_surv(t+x0, T+x0)*(N*BS(t,T,St,pi0/N) + exp(-r*(T-t))*pi0)
  return(reserve)
}
#reserve at t=0 with S0 = 1
V_reserve(t = 0, T = 10, St = 1)
```

## Execise 10.4 
In this exercise we are asked to see what happens to the reserve at time $t=6$, if: 

i) Fund increases by 45% 

ii) Fund increases by 5%

```{r}
#fund has increased by 45%: 
V_reserve(t = 6, T = 10, St = 1.45)

#fund has increased by 5%: 
V_reserve(t = 6, T = 10, St = 1.05)

```




\newpage 
## Exercise 10.5 
We are considering a call option: $X = (S_{T}-K)^{+}$, furthermore the stock price follows a GBM: 
$$
dS_{t} = \mu S_{t}dt + \sigma S_{t}dW_{t}\;\;(P)
$$
We are asked to show that the claim value at time $t$ denoted $v(t,S_{t})$, satisfies the following PDE: 
$$\begin{aligned}
-\partial_{t}v + rv &= 
\frac{1}{2}\sigma^{2}x^{2}\partial_{xx}v + rx\partial_{x}v
\end{aligned}$$

First of, by claim value at time t, we mean: 
$$
v(t,S_{t}) = e^{-r(T-t)}\mathbb{E}_{Q}[X|\mathcal{F}_{t}]
$$
This is an object we can trade, and form the fundamental theorem of asset-pricing we have that all tradable assets are $(Q,\mathcal{F})$-martingales after discounting.
This means that: 
$$
\frac{v(t,S_{t})}{B_{t}}
$$
should be a martingale, here $B_{t} = \exp\left(\int_{0}^{t}r_{u}du\right)$, in our case we have a constant interest rate, meaning that $B_{t} = e^{rt}$.
Our strategy is to use the martigale representation theorem, from which we can conclude that the $dt$-terms should be zero.

$$\begin{aligned}
d\left[\frac{v(t,S_{t})}{B_{t}}\right] &= 
d\left[\frac{1}{B_{t}}\right]v(t,S_{t}) + \frac{1}{B_{t}}dv(t,S_{t}) + d\left[\frac{1}{B_{t}}\right]dv(t,S_{t}) \\ 
&= d\left[\frac{1}{B_{t}}\right]v(t,S_{t}) + \frac{1}{B_{t}}dv(t,S_{t})
\end{aligned}$$

Since we work in $(Q)$-framework, we stat the dynamics of $S_{t}$ under $Q$ first: 
$$
dS_{t} = rS_{t}dt + \sigma S_{t}d\widetilde{W}_{t}\;\; (Q)
$$
We start with the easy one first: 
$$
d(e^{-rt}) = -re^{-rt}dt
$$
We will use Ito's formula on $v(t,x)$:
$$\begin{aligned}
dv(t,x) &= \partial_{t}vdt + \partial_{x}vdS_{t} + \frac{1}{2}\partial_{xx}v(dS_{t})^{2} \\ 
&= \partial_{t}vdt + \partial_{x}v[rxdt + \sigma xd\widetilde{W}_{t}] 
+ \frac{1}{2}\partial_{xx}v\sigma^{2}x^{2}dt \\ 
&= \left[
\partial_{t}v + \partial_{x}vx + \frac{1}{2}\partial_{xx}v\sigma^{2}x^{2}
\right]dt 
+ \partial_{x}vxd\widetilde{W}_{t}
\end{aligned}$$

This leaves us with: 
$$\begin{aligned}
d\left[\frac{v(t,S_{t})}{B_{t}}\right] &= 
-re^{-rt}vdt + e^{-rt}\left(
\left[
\partial_{t}v + \partial_{x}vx + \frac{1}{2}\partial_{xx}v\sigma^{2}x^{2}
\right]dt 
+ \partial_{x}vxd\widetilde{W}_{t}
\right) \\ 
&= 
e^{-rt}\left[
-rv + \partial_{t}v +rx\partial_{x}v + \sigma^{2}x^{2}\frac{1}{2}\partial_{xx}v
\right]dt 
+ e^{-rt}x\partial_{x}vd\widetilde{W}_{t}
\end{aligned}$$

And then, combining the fundamental theorme of asset pricing with martingale representation theorem, we get that the $dt$-part must equal to zero, hence: 

$$\begin{aligned}
-\partial_{t}v + rv &= 
\frac{1}{2}\sigma^{2}x^{2}\partial_{xx}v + rx\partial_{x}v
\end{aligned}$$








