---
title: "Exercises chapter 10"
author: "Andreas Slåttelid"
header-includes: \usepackage{bbm}
output:
  pdf_document:
    extra_dependencies:
    - bbm
    - threeparttable
    - dsfont
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 


## Exercise 10.1 
In this exercise we are dealing with a term-insurance, meaning that $S = \{*,\dagger\}$, we have stochastic interest rates, which follows a Vasicke model, furthermore, the risk-premium is given by $\gamma = 0$, meaning that we will have the same model under $P$ as under $Q$, the contractual information provided is: 
```{r}
T  <- 10 
x0 <- 50 
B  <- 200000

#Vasicek parameters:
a     <- 5 
b     <- 0.04 
sigma <- 0.01 
r0    <- 0.02

#0:alive, 1: dead 
mu01 <- 0.009

p_surv <- function(t,s){
  f <- mu01
  integral <- integrate(f, lower = t, upper = s)$value
  return(exp((-1)*integral))
}
```
We clarify, what the survival-probability is: 
$$\begin{aligned}
p(t,s) &= \exp\left(-\int_{t}^{s}\mu_{*\dagger}(u)du\right) \\ 
&= \exp(-0.009[s-t])
\end{aligned}$$

The desired Vasicek model is in this case given by: 
$$
dr_{t} = a(b-r_{t})dt + \sigma dW_{t}^{Q}
$$


Furthermore it's said that premiums are paid continiously, giving us the following policy-functions: 

$$\begin{aligned}
a_{*}(t) &= 
\begin{cases}
0, &t<0 \\
-\pi t, &t\in[0,T) \\ 
-\pi T, &t\geq T
\end{cases}
\;\;
a_{*\dagger}(t) =
\begin{cases}
B, &t\in[0,T) \\ 
0, &else
\end{cases}
\end{aligned}$$

and 
$$\dot{a}_{*}(t) = -\pi \mathbbm{1}_{[0,T)}(t)$$
We start off by stating the general formula for reserves in continous-time: 
$$\begin{aligned}
V_{i}^{+}(t,A) &= 
\sum_{j}\frac{v(T)}{v(t)}p_{ij}^{x}(t,T)\Delta a_{j}(T) 
+ \sum_{j}\int_{t}^{T}\frac{v(s)}{v(t)}p_{ij}^{x}(t,s)da_{j}(s) 
+ \sum_{k\neq j}\int_{t}^{T}\frac{v(s)}{v(t)}p_{ij}^{x}(t,s)\mu_{jk}^{x}(s)a_{jk}(s)ds
\end{aligned}$$

Furthermore, we define: 
$$
V_{i}^{+}(t,r_{t}) = \mathbb{E}_{Q}[V_{i}^{+}(t,A)|\mathcal{F}_{t}]
$$
we also have: 
$$
P(t,T) = \mathbb{E}_{Q}[e^{-\int_{t}^{T}r_{u}du}|\mathcal{F}_{t}]
$$
We have $i = *$ and $j\in \{*,\dagger\}$, this gives us: 
$$\begin{aligned}
V_{*}^{+}(t,r_{t}) &= -\pi \int_{t}^{T}P(t,s)p_{**}^{x}(t,s)ds 
+ B\int_{t}^{T}P(t,s)p_{**}^{x}(t,s)\mu_{*\dagger}^{x}(s)ds
\end{aligned}$$

Since we are dealing with the Vasicek model, we have that it belongs to a class called: Affine term structures, which in simplicity means that the zero-coupon bond price $P(t,T)$ have an explicit form: 

$$
P(t,T) = e^{-A(T-t)r_{t} + B(T-t)}
$$
Where: 
$$\begin{aligned}
A(h) &= \frac{1-e^{-ah}}{h}\;\; \text{and} \;\; 
B(h) = \left(
b + \frac{\gamma \sigma}{a} - \frac{\sigma^{2}}{2a^{2}}
\right)[A(h)-h] - \frac{\sigma^{2}}{4a}A(h)^{2}
\end{aligned}$$

Now from the equivalence principle, we get that the fair premium $\pi$ is given by:
$$
0 = V_{*}^{+}(0,A) = \mathbb{E}_{Q}[V_{*}^{+}(0,r_{0})]
$$
Hence: 
$$\begin{aligned}
0 &= -\pi \int_{0}^{T}P(0,s)p_{**}^{x}(0,s)ds 
+ B\int_{0}^{T}P(0,s)p_{**}^{x}(0,s)\mu_{*\dagger}^{x}(s)ds \\ 
&\Downarrow \\ 
\pi &= \frac{B\int_{0}^{T}P(0,s)p_{**}^{x}(0,s)\mu_{*\dagger}^{x}(s)ds}{
\int_{0}^{T}P(0,s)p_{**}^{x}(0,s)ds
}
\end{aligned}$$






