---
title: "Exercises chapter 6"
author: "Andreas Slåttelid"
header-includes: \usepackage{bbm}
output:
  pdf_document:
    extra_dependencies:
    - bbm
    - threeparttable
    - dsfont
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 

## Exercise 6.1 
In this exercise we are asked to calculate transition rates using an Euler scheme, we have four states, i.e $S = \{*, \diamond, \times, \dagger\}$, where $\times$ represents critically ill. 
From the text, we have the following transition rate matrix:
```{r, message=FALSE}
#required libraries: 
library(tidyverse)
library(scales)
```


```{r}
#0: alive, 1: disabeld, 2: critcally ill, 3: dead 
lambda <- function(t){
  
  #constants given in exercise:
  a1 <- 4*10**(-4)
  b1 <- 3.4674*10**(-6)
  c1 <- 0.138155 
  a2 <- 5*10**(-4)
  b2 <- 7.5858*10**(-5)
  c2 <- 0.08749
  
  #alive
  m01 <- a1 + b1*exp(c1*t)
  m02 <- 0.05*m01 
  m03 <- a2 + b2*exp(c2*t)
  m00 <- -(m01 + m02 + m03)
  
  #disabeld
  m10 <- 0.1*m01 
  m12 <- m02 
  m13 <- m03 
  m11 <- -(m10 + m12 + m13)
  
  #critically ill
  m23 <- 1.2*m13
  m21 <- 0 
  m20 <- 0 
  m22 <- -(m23 + m21 +m20)
  
  #deceased
  m30 <- m31 <- m32 <- m33 <- 0
  
  transition_rates <- c(m00, m01, m02, m03, 
                        m10, m11, m12, m13, 
                        m20, m21, m22, m23, 
                        m30, m31, m32, m33)
  
  L <- matrix(transition_rates, nrow=4, byrow=TRUE)
  
  return(L)
}
```

```{r, message = FALSE}
field <- function(t,M){
  return(M%*%lambda(t))
}

#field(t0, P0)

Euler <- function(t0,P0,h,tn){
  if(t0==tn){ return(P0)}
  N <- (tn-t0)/h
  D <- dim(P0)[1] #gives dimension of mxn matrix dim = c(m,n)
  #Initial condition at s
  P <- array(diag(D*(N+1)), dim=c(D,D,N+1))
  #First iteration
  P[,,1] <- P0 + h*field(t0,P0)
  
  
  for(n in 1:N){
    P[,,n+1] <- P[,,n]+h*field(t0+n*h,P[,,n])
  }
  return(P) #returns array, array(data , dim= c(3,3,2)), this stores 2 3x3 matricies
}

t0 <- 30       #start age
P0 <- diag(4)  #initial start with P(s,s) = I
tn <- 110      #end age
h <- 1/12      #step size 
N <- (tn-t0)/h #number of steps

sol <- Euler(t0,P0,h,tn) #contains ages transition probs from 30 to 110
#p_surv(30,75): 
sol[1,1, ][12*35]

``` 
\newpage
```{r}
a <- seq(30, 110, length = length(sol[1,1, ]))
b <- sol[1,1, ]


df <- data.frame(a,b)
colnames(df) <- c("age", "survival_prob")

fig <- df %>% 
  ggplot(aes(x = age, y = survival_prob)) + 
  geom_line() + 
  scale_y_continuous() + 
  xlab("Age") + 
  ylab("Survival probability") 

fig
```







\newpage

## Exercise 6.2 

We have the following contractual information: 
```{r}
T <- 10     #length of contract
x <- 60     #age of insured
D <- 20000  #disability pension while disabeld
B <- 50000  #death benefit 
r <- 0.05   #intensity rate
```

Where our state space is $S = \{*, \diamond, \dagger\}$, we have the following policy-functions: 
$$\begin{aligned}
a_{\diamond}(t) &= 
\begin{cases}
0, &t<0 \\
Dt, &t\in[0,T) \\ 
DT, &t\geq T
\end{cases}
\;\;
a_{*\dagger}(t) =
\begin{cases}
B, &t\in[0,T) \\ 
0, &else
\end{cases}
\end{aligned}$$
Since we for now do not care about the premiums in the policy-functions, we get $$\dot{a}_{*}(t) = 0 $$ and $$\dot{a}_{\diamond}(t) = D\mathbbm{1}_{[0,T)}(t)$$


We can set up thieles differential equations: 
$$\begin{aligned}
\frac{d}{dt}V_{*}^{+}(t) &= r(t)V_{*}(t) - \mu_{*\diamond}^{x}(t)[V_{\diamond}^{+}(t) - V_{*}^{+}(t)] - \mu_{*\dagger}^{x}(t)[B\mathbbm{1}_{[0,T)}(t) - V_{*}^{+}(t)] \\ 
\frac{d}{dt}V_{\diamond}^{+}(t) &= r(t)V_{\diamond}(t) -D\mathbbm{1}_{[0,T)}(t) 
- \mu_{*\diamond}^{x}(t)[V_{*}^{+}(t)-V_{\diamond}^{+}(t)] 
+ \mu_{\diamond \dagger}^{x}(t)V_{\diamond}^{+}(t)
\end{aligned}$$
With finial conditions: $V_{*}^{+}(T) = V_{\diamond}^{+}(T) = 0$

In order to solve this numerically, we partition up $[0,T]$ in times $t_{i} = t_{0} + ih$, with $0 = t_{0} < \dots < t_{n} = T$.

We also iterate backwards, giving us: 
$$\begin{aligned}
\frac{V_{*}^{+}(t_{i})- V_{*}^{+}(t_{i-1}) }{h} &= 
rV_{*}(t_{i}) - \mu_{*\diamond}^{x}(t_{i})[V_{\diamond}^{+}(t_{i}) - V_{*}^{+}(t_{i})] - \mu_{*\dagger}^{x}(t_{i})[B - V_{*}^{+}(t_{i})]
\end{aligned}$$

We then solve for $V_{*}(t_{i-1})$, giving us: 
$$\begin{aligned}
V_{*}^{+}(t_{i-1}) &= V_{*}^{+}(t_{i}) - h\left[
rV_{*}^{+}(t_{i}) - \mu_{*\dagger}^{x}(t_{i})[V_{\diamond}^{+}(t_{i})-V_{*}^{+}(t_{i})]
- \mu_{*\dagger}^{x}(t)[B-V_{*}(t_{i})]
\right]
\end{aligned}$$ 

We do the same for $V_{\diamond}^{+}(t_{i-1})$: 

$$\begin{aligned}
V_{\diamond}^{+}(t_{i-1}) &= V_{\diamond}^{+}(t_{i}) - h\left[
rV_{\diamond}^{+}(t_{i})-D\mathbbm{1}_{[0,T)}(t) - \mu_{\diamond *}^{x}(t_{i})[V_{*}^{+}(t_{i})-V_{\diamond}^{+}(t_{i})]
+ \mu_{\diamond \dagger}^{x}(t_{i})V_{\diamond}^{+}(t_{i})
\right]
\end{aligned}$$



```{r}
#states 0:alive, 1:disabled, 2:dead

mu01 <- function(u){
  a1 <- 4*10^{-4}
  b1 <- 3.4674*10^{-6}
  c1 <- 0.138755
  return(a1 + b1*exp(c1*u))
}

mu02 <- function(u){
  a2 <- 5*10^{-4}
  b2 <- 7.5858*10^{-5}
  c2 <- 0.087498
  return(a2 + b2*exp(c2*u))
}

mu12 <- function(u){
  return(mu02(u))
}

mu10 <- function(u){
  return(0.1*mu01(u))
}

```

```{r}
#a.e. derivative of a_{\diamond}(t)
dis_dot <- function(t){
  if(t >= 0 && t<T){ return(D) }
  if(t >= T){ return(0) }
}

#PV of policy V0 and V1
h <- 1/12
N <- T/h

PV_act <- PV_dis <- rep(0,N+1)
PV_act[N+1] <- PV_dis[N+1] <- 0 #by construction of contract

for(i in N:1){
  PV_act[i] <- PV_act[i+1] - h*(r*PV_act[i+1]
                                - mu01(x+(i+1)*h)*(PV_dis[i+1]-PV_act[i+1])
                                - mu02(x+(i+1)*h)*(B-PV_act[i+1])
                                )
  
  PV_dis[i] <- PV_dis[i+1] - h*(r*PV_dis[i+1]- dis_dot((i+1)*h)
                                - mu10(x+(i+1)*h)*(PV_act[i+1]-PV_dis[i+1])
                                + mu12(x+(i+1)*h)*PV_dis[i+1]
                                )
} 
```
\newpage
And from this we can find the one-time premium $\pi_{0} = V_{*}^{+}(0)$: 
```{r}
pi0 <- PV_act[1]
pi0
```
We were asked to find the yearly premium $\pi$, we then use the method, where we create an artificial policy, where one pays a premium of NOK 1 during $[0,T)$, i.e this is the only contribution of interest, thus: 
$$\begin{aligned}
a_{*}^{Prem=1}(t) &= 
\begin{cases}
0, &t<0 \\
-t, &t\in[0,T) \\ 
-T, &t\geq T
\end{cases}
\;\;\implies 
\dot{a}_{*}^{Prem =1}(t) = -1\mathbbm{1}_{[0,T)}(t) \\ 
a_{\diamond}^{Prem=1}(t) &= 0, \; \forall t \;\; \implies \dot{a}_{\diamond}^{Prem = 1} 0 \;\; \\ 
a_{*\dagger}^{Prem=1}(t) &= 0
\end{aligned}$$

We can now set up the new thieles equations: 

$$\begin{aligned}
\frac{d}{dt}V_{*}^{+}(t, A^{Prem=1}) &= rV_{*}^{+}(t) - \dot{a}_{*}(t) - \mu_{*\diamond}^{x}(t)[V_{\diamond}^{+}(t)-V_{*}^{+}(t)]
+\mu_{*\dagger}^{x}(t)V_{*}^{+}(t) \\ 
\frac{d}{dt}V_{\diamond}^{+}(t, A^{Prem=1}) &= rV_{\diamond}(t) 
- \mu_{*\diamond}^{x}(t)[V_{*}^{+}(t)-V_{\diamond}^{+}(t)] 
+ \mu_{\diamond \dagger}^{x}(t)V_{\diamond}^{+}(t)
\end{aligned}$$
And again solve for $V_{*}^{+}(t_{i-1})$ and $V_{\diamond}^{+}(t_{i-1})$: 

$$\begin{aligned}
V_{*}^{+}(t_{i-1}) &= V_{*}^{+}(t_{i}) - h\left[
rV_{*}^{+}(t_{i}) -\dot{a}_{*}(t_{i}) - \mu_{*\dagger}^{x}(t_{i})[V_{\diamond}^{+}(t_{i})-V_{*}^{+}(t_{i})]
+ \mu_{*\dagger}^{x}(t)V_{*}(t_{i})
\right] \\ 
V_{\diamond}^{+}(t_{i-1}) &= V_{\diamond}^{+}(t_{i}) - h\left[
rV_{\diamond}^{+}(t_{i}) - \mu_{\diamond *}^{x}(t_{i})[V_{*}^{+}(t_{i})-V_{\diamond}^{+}(t_{i})]
+ \mu_{\diamond \dagger}^{x}(t_{i})V_{\diamond}^{+}(t_{i})
\right]
\end{aligned}$$ 




```{r}
#a.e derivative a_{*}:
a_dot_prem1 <- function(t){
  if(t >= 0 && t < T){ return(-1) }
  if(t >= T){ return(0) }
}

PV_prem1_act <- PV_prem1_dis <- rep(0,N+1)
PV_prem1_act[N+1] <- PV_prem1_dis[N+1] <- 0 #follows from contract 

for(i in N:1){
  PV_prem1_act[i] <- PV_prem1_act[i+1] - h*(r*PV_prem1_act[i+1] - a_dot_prem1((i+1)*h)
                                - mu01(x+(i+1)*h)*(PV_prem1_dis[i+1]-PV_prem1_act[i+1])
                                + mu02(x+(i+1)*h)*(PV_prem1_act[i+1])
                                )
  
  PV_prem1_dis[i] <- PV_prem1_dis[i+1] - h*(r*PV_prem1_dis[i+1] 
                                - mu10(x+(i+1)*h)*(PV_prem1_act[i+1]-PV_prem1_dis[i+1])
                                + mu12(x+(i+1)*h)*PV_prem1_dis[i+1]
                                )
}
PV_prem1_act[1]
```
We will now use the equvalence principle, which tells us: 
$$\begin{aligned}
\pi V_{*}^{+}(0,A^{Prem=1}) &+ V_{*}^{+}(0) = 0 \\ 
&\Downarrow \\ 
\pi &= -\frac{V_{*}(0)}{V_{*}^{+}(0,A^{Prem=1})}
\end{aligned}$$

This gives us: 
```{r}
pi0

yearly_premium <- (-1)*pi0/PV_prem1_act[1]
yearly_premium
```


## Exercise 6.4 
We are dealing with an endownment insurance, meaning that $S = \{*,\dagger\}$, we are asked to calculate the second moment of the benefit payments $V_{t}^{+}$ for $t = 10, \dots, 20$. 
The contractual information provided is: 
```{r}
x <- 40 
T <- 20 
E <- 115000
B <- 220000
r <- 0.03 
```


We start by writing up the policy functions: 
$$\begin{aligned}
a_{*}^{Pre}(n) &= \begin{cases}
0  & ,n = 0, \dots, T-1 \\
E &,else
\end{cases} \;\; 
a_{*\dagger}^{Post}(n) &= \begin{cases}
B  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases}
\end{aligned}$$


From chapter.5 we are given the following formula: 
$$\begin{aligned}
\mathbb{E}[(V_{t}^{+})^{p}|X_{t}=i]
&= 
(v_{t})^{p}\sum_{j}p_{ij}^{x}(t,t+1)\sum_{k=0}^{p}{p \choose k}(a_{ij}^{Post}(t))^{p-k}\mathbb{E}[(V_{t+1}^{+})^{k}|X_{t+1}=j]
\end{aligned}$$

Where $v_{t}$ is the one-step discount factor in $[t,t+1]$ i.e:
$$
v(t) = v_{0}v_{1}\dots v_{t-1}
$$

Let's denote $V_{2}(t) := \mathbb{E}[(V_{t}^{+})^{2}|X_{t}=*]$, and then translate the equation above into our situation: 

$$\begin{aligned}
V_{2}(t) &= (v_{t})^{2}\sum_{j}p_{*j}^{x}(t,t+1)\sum_{k=0}^{2}{2 \choose k}(a_{*j}^{Post}(t))^{2-k}\mathbb{E}[(V_{t+1}^{+})^{k}|X_{t+1}=j] \\ 
&= (v_{t})^{2}p_{**}^{x}(t,t+1)\mathbb{E}[(V_{t+1}^{+})^{2}|X_{t+1}=*] 
+ (v_{t})^{2}p_{*\dagger}^{x}(t,t+1)(a_{*\dagger}^{Post}(t))^{2} \\ 
&= (v_{t})^{2}p_{**}^{x}(t,t+1)V_{2}(t+1)
+ (v_{t})^{2}p_{*\dagger}^{x}(t,t+1)(a_{*\dagger}^{Post}(t))^{2} \\ 
&= e^{-2r}[p_{**}^{x}(t,t+1)V_{2}(t+1) + p_{*\dagger}^{x}(t,t+1)(a_{*\dagger}^{Post}(t))^{2}]
\end{aligned}$$

This has now become an difference equation, with final condition: 
$$
V_{2}(T) = (\Delta a_{*}^{Pre}(T))^{2} = E^{2}
$$
\newpage 
From the exercise it's also given that: 

$$
\mu_{*\dagger}^{x}(t) = 0.0018 + 0.0004t
$$
```{r}
#0:alive, 1:dead

mu01_x <- function(t){
  return(0.0018 + 0.0004*t)
}

p_surv_x <- function(t,s){
  f <- mu01_x
  integral <- integrate(f, lower = t, upper = s)$value
  return(exp((-1)*integral))
}

```



```{r}
V2 <- rep(0,T+1) 
V2[T+1] <- E^{2}

for(n in (T-1):0){
  V2[n+1] <- exp(-2*r)*(p_surv_x(n,n+1)*V2[n+2] + B^{2}*(1-p_surv_x(n,n+1)))
}
V2
```

\newpage 

## Exercise 6.5 
We are considering a permanent disability insurance, so $S = \{*, \diamond, \dagger \}$, with $\mu_{\diamond *}(t) = 0$. Contractual information: 
```{r}
x <- 60 
T <- 10 
D <- 30000
B <- 50000 
r <- 0.05
u <- 60000
```

We are asked to find: 
$$
P_{\diamond}(t,u) := P[V_{t}^{+} < u|X_{t}=\diamond]
$$
for $t=7$ and $u=\$60k$

Before we start, we write up the desired policy functions: 

$$\begin{aligned}
a_{\diamond}^{Pre}(n) &= \begin{cases}
D  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases} \;\; 
a_{*\dagger}^{Post}(n) &= \begin{cases}
B  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases} \;\; 
a_{\diamond\dagger}^{Post}(n) &= \begin{cases}
B  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases}
\end{aligned}$$ 

We will use Thiele's difference equation for distributions of mathematical reserves: 

$$\begin{aligned}
P_{i}(t,u) &= \sum_{k} p_{ik}^{x}(t,t+1)P_{k}(t+1, v_{t}^{-1}[
u - a_{i}^{Pre}(t)-a_{ik}^{Post}(t)
])
\end{aligned}$$

Translating this into our situation: 
$$\begin{aligned}
P_{\diamond}(t,u) &= p_{\diamond \diamond}^{x}(t,t+1)P_{\diamond}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \diamond}^{Post}(t)
]) \\ 
&+ p_{\diamond \dagger}^{x}(t,t+1)P_{\dagger}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
]) \\ 
&= 
p_{\diamond \diamond}^{x}(t,t+1)P_{\diamond}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)
]) \\ 
&+ p_{\diamond \dagger}^{x}(t,t+1)P_{\dagger}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
])
\end{aligned}$$


We start by exploring $P_{\dagger}$: 
$$\begin{aligned}
P_{\dagger}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
]) &= 
P[V_{t+1}^{+} < e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
]|X_{t+1} = \dagger] \\ 
&= 
P[0 < e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
]] \\ 
&= \begin{cases}
1 &, t>T \\ 
0 &, t\leq T
\end{cases}
\end{aligned}$$
Since we are interested in $t\leq T$, we get: 
$$
P_{\dagger}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)- a_{\diamond \dagger}^{Post}(t)
] = 0
$$
Giving us: 
$$\begin{aligned}
P_{\diamond}(t,u) &= p_{\diamond \diamond}^{x}(t,t+1)P_{\diamond}(t+1,e^{r}[
u - a_{\diamond}^{Pre}(t)
])
\end{aligned}$$

We iterate backwards, given that we know $V_{10}^{+} = 0$, as we do not need to set aside money for someone entering the contract at the end. 
$$\begin{aligned}
P_{\diamond}(9,60) &= p_{\diamond \diamond}^{x}(9,10)P_{\diamond}(10,e^{r}[
60 - a_{\diamond}^{Pre}(9)
]) \\ 
&= p_{\diamond \diamond}^{x}(9,10)P[0 < e^{r}[
60 - 30] \\ 
&= p_{\diamond \diamond}^{x}(9,10)
\end{aligned}$$













