---
title: "Excercises chapter 4 vectorized"
author: "Andreas Slåttelid"
output: 
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

## Exercise 4.4 Term insurance in discrete time 

Required libraries:

```{r, message=FALSE}
#want the map function from purrr:
library(tidyverse)
library(scales)
packageVersion("tidyverse")
```

In this exercise we are dealing with a term insurance in discrete time,
hence: $S = \{*,\dagger\}$, the contractual information provided:

```{r}
x <- 50
T <- 10 
B <- 200000
r <- 0.025
```

We are asked to use the equivalence principle to determine the fair
premium $\pi$, we start of by describing the policy functions:

$$\begin{aligned}
a_{*}^{Pre}(n) &= \begin{cases}
- \pi  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases} \\ 
a_{*\dagger}^{Post}(n) &= \begin{cases}
B  & ,n = 0, \dots, T-1 \\
0 &,else
\end{cases}
\end{aligned}$$

Furthermore the formula for dircrete reserves are given by:
$$\begin{aligned}
V_{i}^{+}(t,A) &= 
\sum_{j}\sum_{n\geq t}\frac{v(n)}{v(t)}p_{ij}^{x}(t,n)a_{j}^{Pre}(n) 
+ \sum_{k \neq j}\sum_{n\geq t}\frac{v(n+1)}{v(t)}p_{ij}^{x}(t,n)p_{jk}^{x}(n,n+1)a_{jk}^{Post}(n)
\end{aligned}$$ In our case, this translates to: $$\begin{aligned}
V_{*}^{+}(t,A) &= 
-\pi\sum_{n=t}^{T-1}\frac{v(n)}{v(t)}p_{**}^{x}(t,n) 
+ B\sum_{n=t}^{T-1}\frac{v(n+1)}{v(t)}p_{**}^{x}(t,n)p_{*\dagger}^{x}(n,n+1)
\end{aligned}$$

The equivalence principle states that we should choose $\pi$ so that
$V_{*}^{+}(0,A) = 0$, giving us:

$$\begin{aligned}
0 &= -\pi\sum_{n=0}^{T-1}v(n)p_{**}^{x}(0,n) 
+ B\sum_{n=0}^{T-1}v(n+1)p_{**}^{x}(0,n)p_{*\dagger}^{x}(n,n+1) \\ 
&\Downarrow \\ 
\pi &= \frac{
B\sum_{n=0}^{T-1}v(n+1)p_{**}^{x}(0,n)p_{*\dagger}^{x}(n,n+1)
}
{
\sum_{n=0}^{T-1}v(n)p_{**}^{x}(0,n)
}
\end{aligned}$$

\newpage

```{r}
#0: alive, 1:dead 
mu01 <- function(u){
  a <- 0.002 
  b <- 0.0005 
  return(a + b*(u-50))
}

p_surv <- function(t,s){
  f <- Vectorize(mu01)
  integral <- integrate(f, lower = t, upper = s)$value
  return(exp((-1)*integral))
}

v <- function(t){
  return(exp(-(r*t)))
}
```

```{r, premium}
upper_summand <- function(n){
  v(n+1)*p_surv(x, x + n)*(1 - p_surv(x + n, x + n + 1))
}

lower_summand <- function(n){
  v(n)*p_surv(x, x + n)
}

premium_yearly <- (B*sum(map_dbl(0:(T-1), upper_summand)))/(sum(map_dbl(0:(T-1),  lower_summand)))

premium_yearly
```

\newpage

## Exercise 4.6 (Endowment policy in discrete time)

Required-libraries

```{r, message=FALSE}
#want the map function from purrr:
library(tidyverse)
library(scales)
packageVersion("tidyverse")
```

Contractual information provided:

```{r, contractual_info}
x <- 35     #age
T <- 25     #length of contract
r <- 0.035  #interest rate 
E <- 125000 #endownment 
B <- 250000 #death-benfit
```

```{r, 4_6_rates_and_prob}
#0: alive, 1: dead
mu01 <- function(t){
  return(0.0015 + 0.0004*(t-35))
}

p_surv <- function(t, s){
  f <- Vectorize(mu01)
  integral <- integrate(f, lower = t, upper = s)$value
  
  ans <- exp(-integral)
  return(ans)
}

#discount factor
v <- function(t){
  return(exp(-(r*t)))
}
```

$$\begin{aligned}
V_{*}^{+}(t,A) &= \frac{v(T)}{v(t)}p_{**}^{x}(t,T)E -\pi\sum_{n=t}^{T-1}\frac{v(n)}{v(t)}p_{**}^{x}(t,n) + B\sum_{n=t}^{T-1}\frac{v(n+1)}{v(t)}p_{**}^{x}(t,n)p_{*\dagger}^{x}(n, n+1)
\end{aligned}$$

We will now use the equivalence principle, which states that the fair
premium $\pi$ should be such that $V_{*}(0,A) = 0$, and using this
method we end up with: $$\begin{aligned}
\pi &= \frac{v(T)p_{**}(x,x+T)E + B\sum_{n=0}^{T-1}v(n+1)p_{**}(x,x+n)p_{*\dagger}(x+n,x+n+1)}{\sum_{n=0}^{T-1}v(n)p_{**}(x,x+n)}
\end{aligned}$$

```{r}
upper_summand <- function(n){
  v(n+1)*p_surv(x, x + n)*(1-p_surv(x + n, x + n +1))
}

lower_summand <- function(n){
  v(n)*p_surv(x, x + n)
}

upper_sum <- sum(map_dbl(0:(T-1), upper_summand))
lower_sum <- sum(map_dbl(0:(T-1), lower_summand))

upper_expression <- v(T)*p_surv(x, x + T)*E + B*upper_sum

yearly_premium <- upper_expression/lower_sum
yearly_premium
```

```{r, reserve}

V_star <- function(t){
  
  #inner summands
  summand_1 <- function(t, n){
    (v(n)/v(t))*p_surv(x + t, x +n)
  }

  summand_2 <- function(t, n){
    (v(n+1)/v(t))*p_surv(t + x, n + x)*(1 - p_surv(x + n, x + n +1))
  }
  
  ans <- (v(T)/v(t))*p_surv(x + t, x + T)*E - 
         yearly_premium*sum(map_dbl(t:(T-1),summand_1, t = t)) + B*sum(
                        map_dbl(t:(T-1), summand_2, t = t))
  
  return(ans)
}

length_contract <- 0:T
reserve <- map_dbl(0:T, V_star)

df <- data.frame(length_contract, reserve)
colnames(df) <- c("length_contract", "reserve")

df %>%
  ggplot(aes(x = length_contract, y = reserve)) + 
  geom_line() + 
  scale_y_continuous(labels = dollar)
```
