---
title: "Exerciseses chapter 4"
author: "Andreas Slåttelid"
date: "23/2/2022"
output: 
  pdf_document:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4.3

In this exercise we are considering a disability insurance, hence $S = \{*,\diamond ,\dagger \}$. The contractual information given are: as follows: 

```{r}
x <- 30     #age
T <- 35     #length of contract
D <- 20000  #disability pension
r <- 0.03   #intereset rate
premium_yearly <- 2500
```

Furthermore the transition rates are given by: 
```{r}
lambda <- function(t){
  #we do not have any t-dependency
  m12 <- 0.0279 
  m13 <- 0.0229
  m11 <- -(m12+m13)
  m21 <- 0
  m23 <- 0.0229
  m22 <- -(m21+m23)
  m31 <- m32 <- m33 <- 0
  L <- matrix(c(m11,m12,m13,m21,m22,m23,m31,m32,m33), nrow=3, byrow=TRUE)
  
  return(L)
}
```

We also have the following policy-functions: 
$$\begin{aligned}
a_{*}(t) &= \begin{cases}
0, &t<0 \\
-\pi t, &t\in[0,T) \\ 
-\pi T, &t\geq T
\end{cases} \\ 
a_{\diamond}(t) &= \begin{cases}
0, &t<0 \\
Dt, &t\in[0,T) \\ 
DT, &t\geq T
\end{cases} 
\end{aligned}$$


We were asked to calculate $V_{j}^{+}(t,A)$ for $j = *, \diamond$ and for $t=30$.
We then get: 
$$\begin{aligned}
V_{*}^{+}(t,A) &= \frac{1}{v(t)}\left[\int_{t}^{T}v(s)p_{**}(x+t,x+s)da_{*}(s)   
+ \int_{t}^{T}v(s)p_{*\diamond}(x+t,x+s)da_{\diamond}(s) 
\right] \\ 
&= \frac{1}{v(t)}\left[-\pi\int_{t}^{T}v(s)p_{**}(x+t,x+s)ds   
+ D\int_{t}^{T}v(s)p_{*\diamond}(x+t,x+s)ds 
\right]
\end{aligned}$$

\newpage 

$$\begin{aligned}
V_{\diamond}^{+}(t,A) &= \frac{1}{v(t)}\left[
\int_{t}^{T}v(s)p_{\diamond \diamond}(x+t, x+s)da_{\diamond}(s)
\right] \\ 
&= \frac{1}{v(t)}\left[D
\int_{t}^{T}v(s)p_{\diamond \diamond}(x+t, x+s)ds
\right] 
\end{aligned}$$
We will now use an Euler-scheme to simulate what's going on, and then use a Riemann-sum to compute the integrals: 
$$
\int_{a}^{b}f(x)dx \approx \sum_{i=0}^{n-1}f(x_{i}^{*})\Delta x
$$
Where $\Delta x = \frac{b-a}{n}$, which in our case corresponds to: $h$ and $x_{i}^{*} \in [x_{i},x_{i+1}]$
```{r}
field <- function(t,M){
  return(M%*%lambda(t))
}

#field(t0, P0)

Euler <- function(t0,P0,h,tn){
  if(t0==tn){ return(P0)}
  N <- (tn-t0)/h
  D <- dim(P0)[1] #gives dimension of mxn matrix dim = c(m,n)
  #Initial condition at s
  P <- array(diag(D*(N+1)), dim=c(D,D,N+1))
  #First iteration
  P[,,1] <- P0
  
  
  for(n in 1:N){
    P[,,n+1] <- P[,,n]+h*field(t0+n*h,P[,,n])
  }
  return(P) #returns array, array(data , dim= c(3,3,2)), this stores 2 3x3 matricies
}

t0 <- 60       #start age
P0 <- diag(3)  #initial start with P(s,s) = I
tn <- 110      #end age
h <- 1/12      #step size 
N <- (tn-t0)/h #number of steps

sol <- Euler(t0,P0,h,tn) #contains ages transition probs from 60 to 65.

v <- function(t){
  return(exp(-r*t))
}


#v(s), but s in [30,35]
v_ages <- sapply(seq(30, 35, by = h), v) #we apply the function v to each element




#sum1 where we do not care about the premium until afterwards, we use Riemann-sum.
s1 <- 0 
for (i in 1:(length(v_ages)-1)){
  s1 <- s1 + v_ages[i]*sol[1,1, ][i]*h #survival i.e * -> *
}

#sum2: do not care about D=20.000 until afterwards
s2 <- 0 
for (i in 1:(length(v_ages)-1)){
  s2 <- s2 + v_ages[i]*sol[1,2,][i]*h #disability i.e * -> dis
}
 
#V*(30, A): 
V_30_A <- (1/v(30))*((-1)*premium_yearly*s1 + D*s2)
V_30_A

#V_[disabeld](30,A): 

s3 <- 0 
for (i in 1:(length(v_ages)-1)){
  s3 <- s3 + v_ages[i]*sol[2,2,][i]*h #disability i.e dis -> dis
}


#V_[disabeld](30,A): 
V_30_dis <- (1/v(30))*D*s3
V_30_dis

```


\newpage 
## Exercise 4.6 
Required-libraries
```{r, message=FALSE}
list.of.packages <- c("tidyverse", "scales")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(tidyverse)
library(scales)   #nice formatting of plots
```

We are dealing with an endownment policy, in discrete time, this means that $S = \{*, \dagger\}$, the contractual information provided is: 
```{r}
x <- 35     #age
T <- 25     #length of contract
r <- 0.035  #interest rate 
E <- 125000 #endownment 
B <- 250000 #death-benfit
```

futhermore the force of mortality is given by: 
```{r}
mu <- function(t){
  return(0.0015 + 0.0004*(t-35))
}
```

discount-factor:
```{r}
v <- function(t){
  return(exp(-r*t))
}
```



We can also calulate the survival probability: 
$$\begin{aligned}
p_{**}(s,t) &= \exp\left(-\int_{s}^{t}\mu_{*\dagger}(u)du\right)
\end{aligned}$$

```{r}
p_surv <- function(s,t){
  integral <- 0.0015*(t-s) + 0.0004*((1/2)*(t^2 -s^2) + 35*(s-t))
  return(exp(-integral))
}
p_surv(35,36)
```

We are asked to calculate the yearly premium $\pi$, but first we start by defining the policy functions specified in the contract.

$$\begin{aligned}
a_{*}^{Pre}(n) &= \begin{cases}
-\pi &, n = 0, \dots, T-1 \\ 
E&, n=T
\end{cases}
\;\;\;
a_{*\dagger}^{Post}(n) &= \begin{cases}
B &, n = 0, \dots, T-1 \\ 
0&, else
\end{cases}
\end{aligned}$$

We can now start to calculate the reserves, but we must be aware of what happens at $n = T$: 

$$\begin{aligned}
V_{*}^{+}(t,A) &= \frac{1}{v(t)}\left[
\sum_{n=t}^{T}v(n)p_{**}(x+t,x+n)a_{*}^{Pre}(n) 
+ \sum_{n=t}^{T-1}v(n+1)p_{**}(x+t, x+n)p_{*\dagger}(x+n, x+n+1)a_{*\dagger}^{Post}(n)
\right] \\ 
&= \frac{1}{v(t)}\left[\alpha + \beta \right]
\end{aligned}$$


Here $\alpha$ corresponds to the first sum, and $\beta$ corresponds to the second sum: 
$$\begin{aligned}
\alpha &= \sum_{n=t}^{T-1}v(n)p_{**}(x+t, x+n)(-\pi) + v(T)p_{**}(x+t, x+T)E \\ 
\beta &= \sum_{n=t}^{T-1}v(n+1)p_{**}(x+t, x+n)p_{*\dagger}(x+n, x+n+1)B
\end{aligned}$$


We will now use the equivalence principle, which states that the fair premium
$\pi$ should be such that $V_{*}(0,A) = 0$, and using this method we end up with: 
$$\begin{aligned}
\pi &= \frac{v(T)p_{**}(x,x+T)E + B\sum_{n=0}^{T-1}v(n+1)p_{**}(x,x+n)p_{*\dagger}(x+n,x+n+1)}{\sum_{n=0}^{T-1}v(n)p_{**}(x,x+n)}
\end{aligned}$$

```{r}
res1 <- 0 
for (n in 0:(T-1)){
  res1 <- res1 + v(n+1)*p_surv(x, x+n)*(1-p_surv(x+n, x+n+1))
}

res2 <- 0 
for (n in 0:(T-1)){
  res2 <- res2 + v(n)*p_surv(x, x+n)
}

above <- B*res1 + v(T)*p_surv(x,x+T)*E #whats above in the premium expression
below <- res2            #whats below in the premium expression

premium_yearly <- above/below
premium_yearly

```
We now also want the prospective reserves: i.e $V_{*}^{+}(t,A)$ for $t = 0, \dots , T$:


```{r}
#Present value of reserve when including premiums. 
V_reserve_1 <- function(x=35, E = 125000 , T=25){
  #x: age of insured
  #B: benefit specified in contract
  #T: length of contract 
  
  V_star_pre <- NULL
  for (t in 0:(T-1)){
    s <- 0 
    for (n in t:(T-1)){
      s <- s + (1/v(t))*(-1)*premium_yearly*v(n)*p_surv(t+x, n+x)
    }
    V_star_pre[t+1] <- s  + (1/v(t))*v(T)*p_surv(x+t, x+T)*E 
  }
  V_star_pre[T+1] <- E #specified by contract 
  return(V_star_pre)
} 
```

\newpage

```{r}

V_reserve_2 <- function(x=35, B=250000, T=25){
  #x: age of insured
  #B: benefit specified in contract
  #T: length of contract 
  
  V_star_post <- NULL
  for (t in 0:(T-1)){
    s <- 0 
    for (n in t:(T-1)){
      s <- s + (1/v(t))*B*v(n+1)*p_surv(x+t, x+n)*(1-p_surv(x+n, x+n+1))
    }
    V_star_post[t+1] <- s 
  }
  V_star_post[T+1] <- 0 #specified by contract
  return(V_star_post)
}

reserve_total <- V_reserve_1() + V_reserve_2()
reserve_total

``` 

```{r, echo=FALSE}
df <- data.frame(seq(0,T,1), reserve_total)
colnames(df) <- c("age_contract", "reserve_total") 
#df

fig1 <- df %>% 
  ggplot(aes(x=age_contract, y = reserve_total)) + 
  geom_line() + 
  scale_y_continuous(labels = dollar) + 
  xlab("Age of contract") + 
  ylab("Total reserve")

fig1

#p1 <- ggplotly(fig1)
#p1
```


