---
title: "Spouse model implemented with K13"
author: "Andreas Slåttelid"
date: 
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spouse model

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(scales)
```

We will be considering a spouse model, the insured will get a yearly pension $P$, in case one of the insured dies. Let insured number one have age $x$ and age of insured number two have age $y$.

The state space will be: $S = \{0,1,2,3\}$

```{r}
x <- 40     #age of insured numb.1
y <- 30     #age of insured numb.2
T <- 80     #length of contract
P <- 50000  #pension
r <- 0.03   #intereset rate
```

The policy functions will look like: $$\begin{aligned}
a_{0}^{Pre}(n) &= \begin{cases}
- \pi  & ,n = 0, \dots, T-1 \\
0 &, else
\end{cases} \\ 
a_{1}^{Pre}(n) = a_{2}^{Pre}(n)  &= \begin{cases}
P  & ,n = 0, \dots, T-1 \\
0  & ,else
\end{cases}
\end{aligned}$$

The formula for the reserve in state $0$ will look like: $$\begin{aligned}
V_{0}^{+}(t,A) &= 
-\pi\sum_{n = t}^{T-1}\frac{v(n)}{v(t)}p_{00}^{(x,y)}(t,n)
+ P\left(
\sum_{n=t}^{T-1}\frac{v(n)}{v(t)}p_{01}^{(x,y)}(t,n) + \sum_{n=t}^{T-1}\frac{v(n)}{v(t)}p_{02}^{(x,y)}(t,n)
\right)
\end{aligned}$$

We assume the individuals to have independent lives, and their sample-spaces are $S_{(i)} = \{*, \dagger\}$, this gives:

$$\begin{aligned}
p_{00}^{(x,y)}(t,n) &= p_{**}^{x}(t,n)p_{**}^{y}(t,n) \\ 
p_{01}^{(x,y)}(t,n) &= p_{*\dagger}^{x}(t,n)p_{**}^{y}(t,n) \\ 
p_{02}^{(x,y)}(t,n) &= p_{**}^{x}(t,n)p_{*\dagger}^{y}(t,n)
\end{aligned}$$

Here I have taken state $1$ to be the state where indivdual aged $x$ dies. Let's assume that we deal with a couple where person aged $x$ is a man, and person aged $y$ is a woman. \newpage

```{r}
#The weights given by finanstilsynet
w <- function(x, G){
  #male: G = "M"
  #female: G = "F"
  #x: age in calender year t,
  if (G == "M"){
    return (min(2.671548-0.172480*x + 0.001485*x**2, 0))
  } 
  else {
    return(min(1.287968-0.101090*x+ 0.000814*x**2,0))
  }
}

mu_kol_2013 <- function(x, G){
  #male
  if (G == "M"){
    return((0.241752+0.004536*10**(0.051*x))/1000)
  }
  #female
  else {
    return((0.085411+0.003114*10**(0.051*x))/1000)
  }
} 

#turning mu into a function of u, so that we can use integrate 
mu <- function(u, x, G, Y = 2022){
  return(mu_kol_2013(x+u, G)*(1 + w(x+u, G)/100)^{(Y+u-2013)})
}

p_surv <- function(x, G, Y, t, s){
  
  if (t == s){
    return(1)
  }
  
  f <- Vectorize(mu)
  integral <- integrate(f, lower = t, upper = s, x=x, Y=Y, G=G)$value
  
  ans <- exp((-1)*integral)
  return(ans)
}

v <- function(t){
  return(exp(-(r*t)))
}
```

Again, by the equivalence principle we determine the yearly premium $\pi$ such that $V_{0}^{+}(0,A) = 0$, this gives:

$$\begin{aligned}
\pi &= \frac{P\left(
\sum_{n = 0}^{T-1}v(n)[p_{01}^{(x,y)}(0,n) + p_{02}^{(x,y)}(0,n)]
\right)}{
\sum_{n=0}^{T-1}v(n)p_{00}^{(x,y)}(0,n)
}
\end{aligned}$$

\newpage

```{r, probs}
#both survive:
p_00 <- function(t, n){
    p_surv(x, G="M", Y=2022, t=t,s = n)*p_surv(y, G="F", Y=2022, t=t, s=n)
  }
  
#man dies, woman survive:
p_01 <- function(t,n){
  (1 - p_surv(x, G="M", Y = 2022, t=t,s = n ))*p_surv(y, G = "F", Y=2022, t=t, s = n)
}
  
#man survive, woman die:
p_02 <- function(t,n){
  p_surv(x, G="M", Y=2022, t=t,s = n )*(1 - p_surv(y, G="F", Y=2022, t=t, s = n))
}

#wife remains alive:
p_11 <- function(t,n){
  p_surv(y, G = "F", Y=2022, t=t, s = n)
} 

#man remains alive: 
p_22 <- function(t,n){
  p_surv(y, G = "M", Y=2022, t=t, s = n)
}
```

```{r, yearly premium}
upper_summand <- function(n){
  prob <- p_01(0,n) + p_02(0,n)
  return(v(n)*prob)
}

lower_summand <- function(n){
  prob <- p_00(0,n)
  return(v(n)*prob)
}

(yearly_premium <- P*sum(map_dbl(0:(T-1), upper_summand))/sum(map_dbl(0:(T-1), lower_summand)))
```

\newpage

```{r, reserve}
#reserve in state 0:
V_0 <- function(t){
  
  summand_1 <- function(t, n){
    (v(n)/v(t))*p_00(t, n)
  }
  
  summand_2 <- function(t, n){
    (v(n)/v(t))*(p_01(t,n) + p_02(t,n))
  }
  
  ans <- (-1)*yearly_premium*sum(map_dbl(t:(T-1),summand_1, t = t)) + P*sum(
                        map_dbl(t:(T-1), summand_2, t = t))
  
  return(ans)
}

#reseve in state 1: 
V_1 <- function(t){
  
  summand <- function(t, n){
    (v(n)/v(t))*p_11(t,n)
  }
  
  ans <- P*sum(map_dbl(t:(T-1), summand, t = t))
  
  return(ans)
}

#reserve in state 2: 
V_2 <- function(t){
  
  summand <- function(t,n){
    (v(n)/v(t))*p_22(t,n)
  }
  
  ans <- P*sum(map_dbl(t:(T-1), summand, t = t))
  
  return(ans)
}
```
\newpage
```{r, plot}
length_contract <- 0:(T-1)
state0 <- map_dbl(length_contract, V_0)


df <- data.frame(length_contract, state0) %>% 
  mutate(state1 = map_dbl(length_contract, V_1)) %>% 
  mutate(state2 = map_dbl(length_contract, V_2)) %>% 
  pivot_longer(!length_contract, names_to = "state", values_to = "reserve")
  
df %>%
  ggplot(aes(x = length_contract, y = reserve)) +
  geom_line(aes(color = state, linetype = state)) +
  scale_y_continuous(labels = dollar) 
```
